---
title: "Hypotheticals vs. counterfactuals"
author: "Tobias Gerstenberg"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::html_document2:
    toc: true
    toc_depth: 4
    theme: cosmo
    highlight: tango
---

# Load packages 

```{r, message=F}
library("RSQLite")
library("tidyjson")
library("knitr")
library("png")
library("grid")
library("kableExtra")
library("DT")
library("lme4")
library("broom.mixed")
library("brms")
library("xtable")
library("emmeans")
library("tidyverse")
```

# Set options 

```{r}
theme_set(theme_classic() + 
    theme(text = element_text(size = 24)))

opts_chunk$set(comment = "",
               fig.show = "hold")

# suppress grouping warning 
options(dplyr.summarise.inform = F)

# set default color scheme in ggplot 
options(ggplot2.discrete.color = RColorBrewer::brewer.pal(9,"Set1"))
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(9,"Set1"))
```

# Helper functions

```{r}
# function for printing out html or latex tables 
print_table = function(data, format = "html", digits = 2){
  if(format == "html"){
    data %>% 
      kable(digits = digits) %>% 
      kable_styling()
  }else if(format == "latex"){
    data %>% 
      xtable(digits = digits,
             caption = "Caption",
             label = "tab:table") %>%
      print(include.rownames = F,
            booktabs = T,
            sanitize.colnames.function = identity,
            caption.placement = "top")
  }
}

# function for RMSE 
rmse = function(x, y){
  sqrt(mean((x - y)^2))
}

# extract means and credible intervals from brms model fit
func_brms_params = function(fit, name){
  fit %>% 
    tidy(effects = "fixed",
         fix.intercept = F,
         conf.method = "HPDinterval") %>% 
    select(term, estimate, contains("conf")) %>% 
    mutate(across(-term, ~ round(., 2)),
           estimate = str_c(estimate, " [", conf.low, ", ", conf.high, "]")) %>% 
    select(term, estimate) %>% 
    pivot_wider(names_from = term,
                values_from = estimate) %>% 
    mutate(name = name) %>% 
    relocate(name)
}
```

# Data 

## Read in data 

Restuls from experiment 1 (causal judgment), experiment 2 (counterfactual judgment) and experiment 3 (hypothetical judgment).  

```{r}
# read in data 
con = dbConnect(SQLite(), dbname = "../../data/experiments_anonymized.db");
df.data = dbReadTable(con, "hypothetical_counterfactual")
dbDisconnect(con)

# filter out incompletes 
df.data = df.data %>% 
  filter(status %in% 3:5) %>%
  filter(codeversion %in% c("experiment_1", "experiment_2", "experiment_3"))

# demographics 
df.demographics = df.data$datastring %>% 
  spread_values(age = jstring("questiondata", "age"),
                gender = jstring("questiondata", "gender"),
                race = jstring("questiondata", "race"),
                ethnicity = jstring("questiondata", "ethnicity"),
                feedback = jstring("questiondata", "feedback")) %>% 
  bind_cols(df.data %>% select(codeversion)) %>% 
  as_tibble() %>% 
  rename(participant = document.id,
         condition = codeversion) %>% 
  mutate(time = difftime(df.data$endhit,
                         df.data$beginhit,
                         units = "mins"),
         condition = factor(condition,
                            levels = str_c("experiment_", 1:3),
                            labels = c("causal", "counterfactual", "hypothetical")),
         age = as.numeric(age))

# main data 
df.long = df.data$datastring %>% 
  as.tbl_json() %>% 
  enter_object("data") %>%
  gather_array("order") %>% 
  enter_object("trialdata") %>% 
  gather_object("index") %>% 
  append_values_string("value") %>% 
  as_tibble() %>% 
  rename(participant = document.id) %>% 
  pivot_wider(names_from = index,
              values_from = value) %>% 
  rename(clip = id,
         outcome = gate_pass,
         rating = response) %>% 
  mutate(across(.cols = c(outcome, rating, replay_times, time),
                .fns = ~ as.numeric(.))) %>% 
  mutate(time = time / 1000) %>% 
  select(participant, clip, order, outcome, replay_times, time, rating) %>% 
  filter(!clip %in% c("p1", "p2")) %>% 
  mutate(clip = factor(clip, levels = 1:8),
         outcome = factor(outcome,
                          levels = c(1, 0),
                          labels = c("ball B went in", "ball B missed"))) %>% 
  arrange(participant, clip) %>% 
  left_join(df.demographics %>% 
              select(participant, condition),
            by = "participant") %>% 
  mutate(condition = factor(condition,
                            levels = c("causal", "hypothetical", "counterfactual")))

# filter participants for whom more than 10 clips were incorrectly shown 
# (5 participants total) 
df.long = df.long %>% 
  group_by(participant) %>% 
  filter(max(order) <= 10) %>% 
  ungroup()
```

## Trial information 

```{r}
df.trialinfo = tibble(clip = 1:8,
                      blocked_initial = c(1, 1, 0, 0, 1, 1, 0, 0),
                      blocked_final = c(0, 1, 1, 0, 0, 1, 1, 0)) %>% 
  mutate(block_moved = (blocked_initial != blocked_final)*1,
         across(everything(), ~ as.factor(.)))
```

## Model predictions

```{r}
# noise simulations 
df.noise_simulations = read.csv("data/grid_search.csv") %>% 
  rename(ball_noise = unoise,
         brick_noise = bnoise) %>% 
  select(-X)

# predictions for best fitting parameters
df.model = read.csv("data/model_predictions.csv") %>% 
  rename(clip = X) %>% 
  mutate(clip = clip + 1,
         clip = as.character(clip),
         across(c(hypothetical, counterfactual),
                ~ . * 100))

df.model.ball_only = read.csv("data/model_predictions_ball_only.csv") %>% 
  rename(clip = X) %>% 
  mutate(clip = clip + 1,
         clip = as.character(clip),
         across(c(hypothetical, counterfactual),
                ~ . * 100))

df.model.brick_only = read.csv("data/model_predictions_brick_only.csv") %>% 
  rename(clip = X) %>% 
  mutate(clip = clip + 1,
         clip = as.character(clip),
         across(c(hypothetical, counterfactual),
                ~ . * 100))
```


# Experiment 1: Hypothetical and counterfactual judgments 

## Stats 

### Demographics 

```{r}
df.tmp = df.long %>% 
  filter(condition != "causal") %>% 
  distinct(participant, condition) %>% 
  left_join(df.demographics,
            by = c("participant", "condition"))
# age & time
df.tmp %>% 
  summarize(age_mean = mean(age),
            age_sd = sd(age),
            n = n(),
            time_mean = mean(time),
            time_sd = sd(time)) %>% 
  mutate(across(contains("age"), ~ round(.)),
         across(contains("time"), ~ round(., 1))) %>% 
  pivot_longer(cols = everything(),
               values_transform = list(value = as.character)) 

# condition
df.tmp %>% 
  count(condition)

# gender
df.tmp %>% 
  count(gender)

# race
df.tmp %>% 
  count(race)

# ethnicity
df.tmp %>% 
  count(ethnicity)
```

### Feedback 

```{r}
df.long %>% 
  filter(condition != "causal") %>% 
  distinct(participant, condition) %>% 
  left_join(df.demographics,
            by = c("participant", "condition")) %>% 
  select(participant, condition, feedback) %>% 
  datatable()
```

### Noise parameter search 

#### Save means from hypothetical and counterfactual condition

```{r, eval=FALSE}
df.means = df.long %>% 
  group_by(condition, clip) %>% 
  summarize(rating = mean(rating)) %>% 
  ungroup()

# write csv with means from hypothetical and counterfactual condition 
df.means %>% 
  filter(condition != "causal") %>% 
  pivot_wider(names_from = condition, 
              values_from = rating) %>% 
  rename(human_cf = counterfactual, 
         human_hp = hypothetical) %>% 
  write.csv(file = str_c("../python/data/", "hpcf_means.csv"))
```

#### Best parameters

```{r}
# model with both parameters
df.param.both = df.noise_simulations %>% 
  filter(loss == min(loss)) %>% 
  mutate(name = "both")

# model with only ball noise
df.param.ball = df.noise_simulations %>% 
  filter(brick_noise == 0) %>% 
  filter(loss == min(loss)) %>% 
  mutate(name = "ball only")

# model with only brick noise
df.param.brick = df.noise_simulations %>% 
  filter(ball_noise == 0) %>% 
  filter(loss == min(loss)) %>% 
  mutate(name = "brick only")

bind_rows(df.param.ball,
          df.param.brick,
          df.param.both) %>% 
  relocate(name) %>% 
  print_table()
```
### Regression 

```{r}
df.regression = df.long %>% 
  filter(condition != "causal") %>% 
  left_join(df.trialinfo,
            by = "clip") %>% 
  mutate(outcome = factor(outcome,
                          levels = c("ball B missed", "ball B went in"),
                          labels = c("miss", "hit")),
         across(contains("blocked"), ~ as.factor(.)),
         across(c(contains("blocked"), outcome), ~ C(., sum))) # sum contrasts
 
fit_hypothetical = brm(formula = rating ~ 1 + blocked_initial + 
                           blocked_final +
                           outcome + 
                           (1 | participant),
     data = df.regression %>% 
       filter(condition == "hypothetical"),
     file = "cache/fit_hypothetical",
     seed = 1)

fit_counterfactual = brm(formula = rating ~ 1 + blocked_initial + 
                           blocked_final +
                           outcome + 
                           (1 | participant),
     data = df.regression %>% 
       filter(condition == "counterfactual"),
     file = "cache/fit_counterfactual",
     seed = 1)
```

### Model comparison

#### Overall

```{r}
df.long %>% 
  filter(condition != "causal") %>% 
  group_by(condition,
           outcome,
           clip) %>% 
  summarize(rating = mean(rating)) %>% 
  ungroup() %>% 
  left_join(df.model %>% 
              pivot_longer(cols = -clip, 
                           names_to = "condition",
                           values_to = "prediction"),
            by = c("clip", "condition")) %>% 
  summarize(r = cor(rating, prediction),
            rs = cor(rating, prediction, method = "spearman"),
            rmse = rmse(rating, prediction)) %>% 
  mutate(across(.fns = ~ round(., 2))) %>% 
  print_table()
```

#### Separated by condition 

```{r}
df.long %>% 
  filter(condition != "causal") %>% 
  group_by(condition,
           outcome,
           clip) %>% 
  summarize(rating = mean(rating)) %>% 
  ungroup() %>% 
  pivot_wider(names_from = condition,
              values_from = rating) %>% 
  rename_with(.cols = c(hypothetical, counterfactual),
              .fn = ~ str_c(., "_rating")) %>% 
  left_join(df.model %>% 
              rename_with(.cols = c(hypothetical, counterfactual),
              .fn = ~ str_c(., "_model")),
            by = "clip") %>% 
  summarize(r.hypothetical = cor(hypothetical_rating, hypothetical_model),
            r.counterfactual = cor(counterfactual_rating, counterfactual_model),
            rs.hypothetical = cor(hypothetical_rating, hypothetical_model,
                                  method = "spearman"),
            rs.counterfactual = cor(counterfactual_rating, counterfactual_model,
                                    method = "spearman"),
            rmse.hypothetical = rmse(hypothetical_rating, hypothetical_model),
            rmse.counterfactual = rmse(counterfactual_rating, counterfactual_model)) %>% 
  mutate(across(.fns = ~ round(., 2))) %>% 
  pivot_longer(cols = everything(), 
               names_to = c("measure", "comparison"),
               names_sep = "\\.") %>% 
  pivot_wider(names_from = measure,
              values_from = value) %>% 
  print_table()
```

## Plots 

### Hypothetical and counterfactual judgments vs. model predictions 

```{r fig.height=6, fig.width=8}
set.seed(1)

df.plot = df.long %>% 
  filter(condition %in% c("counterfactual", "hypothetical")) %>% 
  mutate(outcome = factor(outcome, levels = c("ball B went in", "ball B missed")))

df.plot2 = df.model %>%
# df.plot2 = df.model.ball_only %>% 
# df.plot2 = df.model.brick_only %>% 
  pivot_longer(cols = -clip,
               names_to = "condition",
               values_to = "rating") %>% 
  mutate(outcome = ifelse(clip <= 4, "ball B went in", "ball B missed"),
         outcome = factor(outcome, levels = c("ball B went in","ball B missed")),
         condition = factor(condition, levels = c("hypothetical", "counterfactual"))) %>% 
  mutate(clip = factor(clip))

p = ggplot(data = df.plot,
           mapping = aes(x = clip, 
                         y = rating,
                         fill = condition)) +
  stat_summary(fun = "mean",
               geom = "bar",
               position = position_dodge(width = 0.9),
               color = "black",
               alpha = 0.5) + 
  stat_summary(fun.data = "mean_cl_boot",
               geom = "linerange",
               position = position_dodge(width = 0.9),
               size = 1.5) + 
  geom_point(data = df.plot2,
             shape = 21,
             position = position_dodge(0.9),
             size = 4,
             show.legend = F) + 
  facet_grid(cols = vars(outcome),
             scales = "free") +
  ylab("mean rating") + 
  scale_y_continuous(breaks = seq(0, 100, 25),
                     labels = seq(0, 100, 25),
                     expand = c(0, 0)) + 
  scale_fill_brewer(palette = "Set1",
                    direction = -1) + 
  coord_cartesian(clip = "off",
                  ylim = c(0, 100)) + 
  theme(legend.position = "bottom",
        legend.margin = margin(t = 2.8, unit = "cm"),
        legend.background = element_blank(),
        axis.title.x = element_blank()) 
    
  
# add images of bricks as x-axis labels  
g = list()
image_names = c("down_up", "down", "up_down", "up")

for(i in 1:length(image_names)){
  img = readPNG(str_c("../../figures/diagrams/blocks/", image_names[i], ".png"))
  g[[i]] = rasterGrob(img, interpolate = TRUE)
  
  p = p +
    annotation_custom(grob = g[[i]],
                      xmin = i - 1,
                      xmax = i + 1,
                      ymin = -40,
                      ymax = -10)
}
print(p)

ggsave(filename = str_c("../../figures/plots/", 
                        "hypothetical_counterfactual_judgments.pdf"),
       plot = p,
       width = 8,
       height = 6)
```

### Noise parameters

```{r}
ggplot(data = df.noise_simulations,
       mappin = aes(x = ball_noise, y = brick_noise)) +
  geom_tile(aes(fill = loss),
            color = "black") +
  scale_fill_gradient(low = "white", high = "black") + 
  coord_cartesian(expand = F) +
  scale_x_continuous(breaks = seq(0, 1.2, 0.2)) + 
  labs(x = expression("ball motion noise"~sigma["ball"]),
       y = expression("block movement noise"~~sigma["block"]))
  
ggsave(filename = "../../figures/plots/parameter_search.pdf",
       width = 8,
       height = 6)
```

## Tables

```{r, warning=F}
func_brms_params(fit_hypothetical, "hypothetical") %>% 
  bind_rows(func_brms_params(fit_counterfactual, "counterfactual")) %>% 
  rename(intercept = Intercept,
         block_initial = blocked_initial1,
         block_final = blocked_final1,
         outcome = outcome1) %>% 
  print_table()
```

# Experiment 2: Causal judgments 

## Stats 

### Demographics 

```{r}
df.tmp = df.long %>% 
  filter(condition == "causal") %>% 
  distinct(participant, condition) %>% 
  left_join(df.demographics,
            by = c("participant", "condition"))

# condition
df.tmp %>% 
  count(condition)

# age & time
df.tmp %>% 
  summarize(age_mean = mean(age),
            age_sd = sd(age),
            n = n(),
            time_mean = mean(time),
            time_sd = sd(time)) %>% 
  mutate(across(contains("age"), ~ round(.)),
         across(contains("time"), ~ round(., 1))) %>% 
  pivot_longer(cols = everything(),
               values_transform = list(value = as.character)) 

# gender
df.tmp %>% 
  count(gender)

# ethnicity
df.tmp %>% 
  count(ethnicity)
```

### Feedback 

```{r}
df.long %>% 
  filter(condition == "causal") %>% 
  distinct(participant, condition) %>% 
  left_join(df.demographics,
            by = c("participant", "condition")) %>% 
  select(participant, condition, feedback) %>% 
  datatable()
```

### Regression 

```{r}
df.regression = df.long %>% 
  filter(condition == "causal") %>% 
  left_join(df.trialinfo,
            by = "clip") %>% 
  mutate(outcome = factor(outcome,
                          levels = c("ball B missed", "ball B went in"),
                          labels = c("miss", "hit"))) %>% 
  mutate(across(contains("blocked"), ~ as.factor(.)),
         across(c(contains("blocked"), outcome), ~ C(., sum))) # setting sum contrasts
 
fit_causal = brm(formula = rating ~ 1 + (blocked_initial + 
                           blocked_final) *
                           outcome + 
                           (1 | participant),
     data = df.regression,
     file = "cache/fit_causal",
     seed = 1)

fit_causal
```

### Model comparison

```{r}
df.means = df.long %>% 
  group_by(condition,
           outcome,
           clip) %>% 
  summarize(rating = mean(rating)) %>% 
  ungroup() %>% 
  mutate(rating = ifelse(condition != "causal" & 
                             outcome == "ball B went in",
                         100 - rating,
                         rating)) %>% 
  pivot_wider(names_from = condition,
              values_from = rating)

df.means %>% 
  summarize(r.causal_hypothetical = cor(causal, hypothetical),
            r.causal_counterfactual = cor(causal, counterfactual),
            rs.causal_hypothetical = cor(causal, hypothetical, method = "spearman"),
            rs.causal_counterfactual = cor(causal, counterfactual, method = "spearman"),
            rmse.causal_hypothetical = rmse(causal, hypothetical),
            rmse.causal_counterfactual = rmse(causal, counterfactual)) %>% 
  mutate(across(.fns = ~ round(., 2))) %>% 
  pivot_longer(cols = everything(), 
               names_to = c("measure", "comparison"),
               names_sep = "\\.") %>% 
  pivot_wider(names_from = measure,
              values_from = value) %>% 
  print_table()
```

## Plots 

### Causal judgments

```{r fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
set.seed(1)

df.plot = df.long %>% 
  filter(condition == "causal") %>% 
  mutate(outcome = factor(outcome, levels = c("ball B went in", "ball B missed")))

p = ggplot(data = df.plot,
           mapping = aes(x = clip,
                         y = rating)) +
  stat_summary(fun = "mean",
               geom = "bar",
               color = "black",
               fill = "gray90") +
  stat_summary(fun.data = "mean_cl_boot",
               geom = "linerange",
               color = "black",
               size = 1.5) +
  stat_summary(data = df.long %>% 
                 filter(condition == "hypothetical") %>% 
                 mutate(rating = ifelse(outcome == "ball B went in", 100 - rating, rating)),
               fun.data = "mean_cl_boot",
               geom = "pointrange",
               aes(fill = "hypothetical"),
               position = position_nudge(x = -0.2),
               shape = 21,
               size = 1,
               fatten = 5,
               stroke = 1) + 
  stat_summary(data = df.long %>%
                 filter(condition == "counterfactual") %>%
                 mutate(rating = ifelse(outcome == "ball B went in", 100 - rating, rating)),
               fun.data = "mean_cl_boot",
               geom = "pointrange",
               aes(fill = "counterfactual"),
               position = position_nudge(x = 0.2),
               shape = 21,
               size = 1,
               fatten = 5,
               stroke = 1) +
  facet_grid(cols = vars(outcome),
             scales = "free") +
  labs(y = "mean causal rating",
       fill = "model prediction") + 
  scale_y_continuous(breaks = seq(0, 100, 25),
                     labels = seq(0, 100, 25),
                     expand = c(0, 0)) + 
  scale_fill_manual(values = c("hypothetical" = "#377eb8",
                               "counterfactual" = "#e41a1c")) +
  coord_cartesian(clip = "off",
                  ylim = c(0, 100)) + 
  theme(legend.position = "bottom",
        legend.margin = margin(t = 2.8, unit = "cm"),
        legend.background = element_blank(),
        axis.title.x = element_blank()) 

# add images of bricks as x-axis labels  
g = list()
image_names = c("down_up", "down", "up_down", "up")

for(i in 1:length(image_names)){
  img = readPNG(str_c("../../figures/diagrams/blocks/", image_names[i], ".png"))
  g[[i]] = rasterGrob(img, interpolate = TRUE)
  
  p = p +
    annotation_custom(grob = g[[i]],
                      xmin = i - 1,
                      xmax = i + 1,
                      ymin = -40,
                      ymax = -10)
}

print(p)

ggsave(filename = "../../figures/plots/causal_ratings_all.pdf",
       plot = p,
       width = 8,
       height = 6)
```

## Tables

### brms

```{r, warning=F}
func_brms_params(fit_causal, "causal") %>% 
  rename(intercept = Intercept,
         block_initial = blocked_initial1,
         block_final = blocked_final1,
         outcome = outcome1,
         `block_initial:outcome` = `blocked_initial1:outcome1`,
         `block_final:outcome` = `blocked_final1:outcome1`) %>% 
  print_table()
```

# Session info

```{r, echo=F}
sessionInfo()
```